Liveness-Based Garbage Collection for Lazy Languages

### 2015-07-07 paper/paper.tex
Correctness proof

\subsection{Liveness Analysis for lazy languages}

Figure~\ref{fig:live-judge}  describes  our  analysis  which  has  two
parts. The function $\mathit{ref}$,  takes  an application $s$ and
a demand $\sigma$, and returns  the incremental liveness generated for the
free  variables   of  $s$  due  to  the   application.   The  function
$\mathcal{L}$  uses   $\mathit{ref}$  to  propagate   liveness  across
expressions.

@@

We shall now  outline a proof of correctness  of the liveness analysis
described earlier.   This requires  a significant modification  to the
technique  used  in  \cite{asati14lgc}.   We shall  first  modify  the
standard semantics. This semantics, which we shall call {\em minefield
  semantics}, will  model a liveness based garbage  collection at each
transition. This is done by starting from the root-set and inserting a
special value  $\bot$ at each reference  location in the  heap that is
reachable but not live by  our analysis.  Attempts to dereference such
locations  during the  transition will  result in  entering  a special
state denoted  \bang.  We shall then  show that no  program enters the
\bang\ state.

To set up the minefield semantics, we follow these steps:
\begin{enumerate}
\item Given an  demand transformer \Lfonly\  we enrich  the abstract machine
  state  $\rho, S,  H, e$  to $\rho,  S, H,  e, \sigma,  \Sigma$. Here
  $\sigma$ is  the demand on  the currently active function,  thus the
  liveness $\Lv$   at $e$ is $\mathcal{L}(e_f, \sigma, \Lfonly)$, and
  $\Sigma$ is a stack of  demands---one for each function frame pushed
  in $S$.   Second, we  augment $Val$ with  a value $\bot$.   To model
  strong  liveness  $\bot$ may  be  copied  freely,  including into  a
  \CONS\ cell, but explodes  when used computationally (in a primitive
  operation other  than a copy). Additionally we  define $GC (\Lv,
  \Sigma) : (\rho, H, S) \rightarrow (\rho , H , S )$ which determines
  live-reachability using $\rho$  and the $\rho$'s in $S$  as the root
  set  and  following  links  in   $H$  only  as  far  as  allowed  by
  \Lv\  and  $\Sigma$.   Hence  $GC  (\Lv,  \Sigma)$  replaces
  live-unreachable  values---in  $\rho$,  in  $H$ and  the  $\rho$'s  in
  $S$---with  $\bot$.  For  example,  if $x.\epsilon  \in \Lv$  then
  $\rho (x) = \bot$.  Only $GC (\cdots )$ introduces $\bot$. 
\end{enumerate}

@@

From Sections \ref{sec:liveness}  and \ref{sec:computing}, it is clear
that the results of liveness analysis of any program can be modeled by
the kind  of grammar described above. The  following proposition shows
that the converse also holds.
\begin{proposition}
Given    a     grammar    $G$    of    the     form    described    in
Definition~\ref{def:specialgrammar},  it is  possible  to construct  a
program $p$  with program points  $\pi_i$ and variables $x$  such that
the liveness  analysis of $p$  is same as  $G$ except for a  change in
non-terminal names which now have the form \var{\Lanv{i}{x}}.
\end{proposition}
We now show that the decision problem required at garbage collection
time is undecidable.
\newcommand{\state}{\ensuremath{\mathsf{S}}}
\newcommand{\nont}[2]{\ensuremath{\mathsf{S}_{#1}^{#2}}}
\cblue{
\begin{lemma}
Consider    a    grammar    $G$    of   the    kind    described    in
Definition~\ref{def:specialgrammar} and a forward access path $\alpha$
consisting  of symbols \acar\  and \acdr\  only. The  decision problem
$\alpha \in \mathscr{L}(G)$ is undecidable.
\end{lemma} 
\begin{proof}
Given a  Turing machine  and an input  $w\in (1+0)^*$, we  construct a
grammar $G$  such that  the machine will  halt on  $w$ if and  only if
$\epsilon \in  \mathscr{L}(G)$. The grammar includes the  fixed set of
unrestricted productions  in Definition~\ref{def:specialgrammar}.

@@

Assume that  the TM starts  in a  state $S_\mathit{init}$ with  a tape
$cw$ and  the head positioned on  the symbol $c$. Then  the sentential
form    corresponding     to    the    initial     configuration    is
$\nont{\mathit{init}}{c}w$ (we  can assume that there  is a production
$\nont{}{} \rightarrow  \nont{\mathit{init}}{c}w$, where  \nont{}{} is
the start symbol of the  grammar). Further correspondences between the
Turing machine moves and the grammar productions are as follows:

@@

After  the Turing machine  has halted,  there are  further ``cleanup''
derivations  that derive  $\epsilon$ only  if the  invariant  has been
maintained  so far.  For every  symbol  $c$, we introduce  a  non-terminal
$\nont{\mathit{final}}{c}$ where $\nont{\mathit{final}}{}$ is the
final state of the Turing machine. We add
productions $\nont{\mathit{final}}{c} \rightarrow \acar
\nont{\mathit{final}}{c}$ and $\nont{\mathit{final}}{c} \rightarrow \acdr
\nont{\mathit{final}}{c}$ for cleaning up the \bcar\ and \bcdr\ symbols on
the left of the head and $\nont{\mathit{final}}{c} \rightarrow 
\nont{\mathit{final}}{c}\bcar$ and $\nont{\mathit{final}}{c} \rightarrow \acdr
\nont{\mathit{final}}{c}\bcdr$ for cleaning up \acar\ and \acdr\ on the
right of the tape head. This completes the reduction.  
\end{proof}
}%end of cblue
We circumvent the problem  of undecidability by over- approximating the
CFG by non-deterministic finite state automata (NFA). The NFA are then
simplified using  the $\hookrightarrow$ rules.  Finally the simplified
NFAs are converted to DFAs.

@@

\begin{figure*}[t!]
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
Premise & Transition & Rule name \\
\hline
\hline
          &\makecell{ $\rho, (\rho', x, e, \{\psi_{i}\}, \sigma')\!:\!S, H, \kappa, \sigma$ \\ $\rightsquigarrow \rho', S, H[\rho'(x) := \kappa], e, \sigma'$ }   &  \sc{const}
\\
\hline
          & \makecell{$\rho, (\rho', z, e, \{\psi_{i}\}, \sigma')\!:\!S, H, (\CONS~x~y), \sigma$ \\ $\rightsquigarrow  \rho', S, H[z := (H(\rho(x)),H(\rho(y)))], e, \sigma'$}     &  \sc{cons} \\
\hline
$H(\rho(x)) \mbox{ is } (d_1, d_2)$ & \makecell{$\rho, (\rho', z, e, \{\psi_{i}\}, \sigma' )\!:\!S, H, (\CAR~x), \sigma$ \\ $ \rightsquigarrow \rho', S, H[\rho(z) := d_1], e, \sigma'$}      &
\sc{car-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{ $\rho, S, H, (\CAR~x), \sigma$ \\ $\rightsquigarrow \rho', (\rho, x, (\CAR~x), \{\psi_{i}\}, \sigma)\!:\!S, H, s,\bigcup_{\psi \in \{\psi_{i}\}}\mmap{}{\psi}{x} $}      &
\sc{car-clo}
\\
\hline
$H(\rho(x)), H(\rho(y)) \in \mathbb{N}$
 & \makecell{$\rho, (\rho', z, e, \{\psi_{i}\}, \sigma')\!:\!S, H, (+~x~y), \sigma$ \\  $\rightsquigarrow \rho', S, H[\rho(z) \mapsto H(\rho(x)) + H(\rho(y))], e, \sigma'$}      &
\sc{prim-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{$\rho, S, H, (+~x~y), \sigma$ \\ $\rightsquigarrow \rho', (\rho, x, (+~x~y), \{\psi_{i}\}, \sigma)\!:\!S, H, s, \bigcup_{\psi \in \{\psi_{i}\}}\mmap{}{\psi}{x}$}      &
\sc{prim-1-clo} \\
\hline
$H(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & \makecell{$\rho, S, H, (+~x~y), \sigma$ \\ $\rightsquigarrow \rho', (\rho, y, (+~x~y), \{\psi_{i}\}, \sigma)\!:\!S, H, s, \bigcup_{\psi \in \{\psi_{i}\}}\mmap{}{\psi}{y} $}      &
\sc{prim-2-clo} \\
\hline
{$\mathit{f}~\mbox{defined as}$
$~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & \makecell{$\rho, S, H, (f~\myvec{x}), \sigma$ \\ $\rightsquigarrow [\myvec{y} \mapsto \rho(\myvec{x})], S, H, e_{\mathit{f}}, \sigma$}      &
\sc{funcall} \\
\hline
$\ell$ is a new location& \makecell{$\rho, S, H, (\LET~x\leftarrow s~\IN~e)$ \\ $ \rightsquigarrow \rho\oplus[x \mapsto \ell], S, H[\ell \mapsto \langle s, \lfloor\rho\rfloor_{FV(s)}  \oplus [x \mapsto
  \ell]\rangle], e$} &
\sc{let} \\
\hline
$H(\rho(x)) \ne 0$ & \makecell{$\rho, S, H, (\pi:\SIF~\psi:x~e_1~e_2), ep, \sigma$ \\ $\rightsquigarrow \rho, S, H,  e_1, ep, \sigma$} & \sc{if-true} \\
\hline
$H(\rho(x)) = 0$ & \makecell{$\rho, S, H, (\pi:\SIF~\psi:x~e_1~e_2), ep, \sigma$ \\   $\rightsquigarrow
\rho, S, H,  e_2, ep, \sigma$} & \sc{if-false} \\
\hline
$H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H,  (\pi:\SIF~\psi:x~e_1~e_2), \sigma$ \\ $\rightsquigarrow
\rho', (\rho, x, (\SIF~x~e_1~e_2),  \pmap{}(\pi), \sigma)\!:\!S, H,  s, \mmap{}{\psi}{x}$}
&
\sc{if-clo} \\
\hline
{$H(\rho(x))~\mbox{is}$ $\mbox{whnf with value}~v$}& \makecell{$\rho, (\rho', z, e, ep, \sigma')\!:\!S, H, (\SRETURN~x), \sigma$ \\ $\rightsquigarrow \rho', S, H[\rho(z) \mapsto v], e, \sigma'$} &
\sc{return-whnf}\\
\hline
$H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H, (\psi:\SRETURN~x), \sigma$ \\ $
  \rightsquigarrow$
$\rho',~ (\rho, x, (\SRETURN~x), \{\psi\}, \sigma)\!:\!S, H,  s, \mmap{}{\psi}{x}$} &
\sc{return-clo} \\
\hline
\end{tabular}
\caption{Minefield semantics for the language.\label{fig:minefield-semantics}}
\end{center}
\end{figure*}

>>>>

\subsection{Liveness Analysis for lazy languages}
\label{sec:liveness-analysis}
Figure~\ref{fig:live-judge}  describes  our  analysis  which  has  two
parts. The function $\mathit{ref}$,  takes  an application $s$ and
a demand $\sigma$, and returns  the incremental liveness generated for the
free  variables   of  $s$  due  to  the   application.   The  function
$\mathcal{L}$  uses   $\mathit{ref}$  to  propagate   liveness  across
expressions.

@@

We shall now  outline a proof of correctness  of the liveness analysis
presented  in  Section~\ref{sec:liveness-analysis}.   This requires  a
significant modification  to the technique  used in \cite{asati14lgc}.
We shall first modify the standard semantics. This semantics, which we
shall  call {\em  minefield semantics},  will model  a  liveness based
garbage collection at  each transition. This is done  by starting from
the root-set  and inserting a  special value $\bot$ at  each reference
location in the  heap that is reachable but not  live by our analysis.
Attempts  to dereference  such  locations during  the transition  will
result in entering a special  state denoted \bang.  We shall then show
that no program enters the \bang\ state.

To set up the minefield semantics, we follow these steps:
\begin{enumerate}
\item  Given an  demand transformer  \Lfonly\ we  enrich  the abstract
  machine state $\rho, S, H, e$  to $\rho, S, H, e, \sigma$.  $\sigma$
  is the  demand on the  currently active function. As  a consequence,
  the liveness $\Lv$ at $e$ is $\mathcal{L}(e, \sigma, \Lfonly)$. Each
  entry  in   the  stack  $S$   of  suspended  evaluations,   is  also
  augmented---it is  now $(\rho, x,  e, eps, \sigma)$. The $\sigma$  in the
  stack entry  represents the demand  on the  function where  the evaluation
  \ldots.  And $eps$ is \ldots.
\item We introduce a function $GC$ that takes the 4-tuple $(\rho, H, S,
  \sigma)$ and  inserts $\bot$ in all locations in the current
  environment $\rho$, the heap $H$ and the environments in the stack
  $S$ that are dead. More formally:

and $\Sigma$ is a  stack of demands---one for each function
  frame pushed in $S$.  Second,  we augment $Val$ with a value $\bot$.
  To model strong liveness $\bot$ may be copied freely, including into
  a  \CONS\  cell,  but  explodes  when  used  computationally  (in  a
  primitive operation  other than a copy). Additionally  we define $GC
  (\Lv, \Sigma)  : (\rho,  H, S) \rightarrow  (\rho ,  H , S  )$ which
  determines live-reachability using $\rho$ and the $\rho$'s in $S$ as
  the root  set and following links in  $H$ only as far  as allowed by
  \Lv\   and   $\Sigma$.     Hence   $GC   (\Lv,   \Sigma)$   replaces
  live-unreachable  values---in $\rho$,  in  $H$ and  the $\rho$'s  in
  $S$---with $\bot$.  For example,  if $x.\epsilon \in \Lv$ then $\rho
  (x) = \bot$.  Only $GC (\cdots )$ introduces $\bot$.
\end{enumerate}

@@

From Sections \ref{sec:liveness}  and \ref{sec:computing}, it is clear
that the results of liveness analysis of any program can be modeled by
the kind  of grammar described above. The  following proposition shows
that the converse also holds.
\begin{proposition}
Given    a     grammar    $G$    of    the     form    described    in
Definition~\ref{def:specialgrammar},  it is  possible  to construct  a
program $p$  with program points  $\pi_i$ and variables $x$  such that
the liveness  analysis of $p$  is same as  $G$ except for a  change in
non-terminal names which now have the form \var{\Lanv{i}{x}}.
\end{proposition}
We now show that the decision  problem that is required to be answered
at       garbage      collection       time       is      undecidable.
\newcommand{\state}{\ensuremath{\mathsf{S}}}
\newcommand{\nont}[2]{\ensuremath{\mathsf{S}_{#1}^{#2}}}  
\begin{lemma}
Consider    a    grammar    $G$    of   the    kind    described    in
Definition~\ref{def:specialgrammar} and a forward access path $\alpha$
consisting  of symbols \acar\  and \acdr\  only. The  decision problem
$\alpha \in \mathscr{L}(G)$ is undecidable.
\end{lemma} 
\begin{proof}
Given a  Turing machine  and an input  $w\in (1+0)^*$, we  construct a
grammar $G$  such that  the machine will  halt on  $w$ if and  only if
$\epsilon \in  \mathscr{L}(G)$. The grammar includes the  fixed set of
unrestricted productions  in Definition~\ref{def:specialgrammar}.

@@

Assume that  the TM  starts in a  state $S_\mathit{init}$ with  a tape
$cw$  and the  head   positioned on  the symbol  $c$. Then  the the
sentential  form   corresponding  to  the   initial  configuration  is
$\nont{\mathit{init}}{c}w$ (we can assume  that there is a production
$\nont{}{}  \rightarrow \nont{\mathit{init}}{c}w$, where  \nont{}{} is
the start symbol of  the grammar). Further correspondences between the
Turing machine moves and the grammar productions are as follows:

@@

After  the Turing machine  has halted,  there are  further ``cleanup''
derivations  that derive  $\epsilon$ only  if the  invariant  has been
maintained  so far.  For every  symbol  $c$, we introduce  a  non-terminal
$\nont{\mathit{final}}{c}$ where $\nont{\mathit{final}}{}$ is the
final state of the Turing machine. We add
productions $\nont{\mathit{final}}{c} \rightarrow \acar
\nont{\mathit{final}}{c}$ and $\nont{\mathit{final}}{c} \rightarrow \acdr
\nont{\mathit{final}}{c}$ for cleaning up the \bcar\ and \bcdr\ symbols on
the left of the head and $\nont{\mathit{final}}{c} \rightarrow 
\nont{\mathit{final}}{c}\bcar$ and $\nont{\mathit{final}}{c} \rightarrow 
\nont{\mathit{final}}{c}\bcdr$ for cleaning up \acar\ and \acdr\ on the
right of the tape head. This completes the reduction.
\end{proof}

We circumvent the problem  of undecidability by over- approximating the
CFG by non-deterministic finite state automata (NFA). The NFA are then
simplified using  the $\hookrightarrow$ rules.  Finally the simplified
NFAs are converted to DFAs.

@@

\begin{figure*}[h!]
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
Premise & Transition & Rule name \\
\hline
\hline
          &\makecell{ $\rho, (\rho', x, e, \{\psi_{i}\}, \sigma')\!:\!S, H, \kappa, \sigma$ \\ $\rightsquigarrow \rho', S, H[\rho'(x) := \kappa], e, \sigma'$ }   &  \sc{const}
\\
\hline
          & \makecell{$\rho, (\rho', z, e, \{\psi_{i}\}, \sigma')\!:\!S, H, (\CONS~x~y), \sigma$ \\ $\rightsquigarrow  \rho', S, H[z := (H(\rho(x)),H(\rho(y)))], e, \sigma'$}     &  \sc{cons} \\
\hline
$H(\rho(x)) \mbox{ is } (d_1, d_2)$ & \makecell{$\rho, (\rho', z, e, \{\psi_{i}\}, \sigma' )\!:\!S, H, (\CAR~x), \sigma$ \\ $ \rightsquigarrow \rho', S, H[\rho(z) := d_1], e, \sigma'$}      &
\sc{car-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{ $\rho, S, H, (\CAR~x), \sigma$ \\ $\rightsquigarrow \rho', (\rho, x, (\CAR~x), \{\psi_{i}\}, \sigma)\!:\!S, H, s,\bigcup_{\psi \in \{\psi_{i}\}}\mmap{}{\psi}{x} $}      &
\sc{car-clo}
\\
\hline
$H(\rho(x)), H(\rho(y)) \in \mathbb{N}$
 & \makecell{$\rho, (\rho', z, e, \{\psi_{i}\}, \sigma')\!:\!S, H, (+~x~y), \sigma$ \\  $\rightsquigarrow \rho', S, H[\rho(z) \mapsto H(\rho(x)) + H(\rho(y))], e, \sigma'$}      &
\sc{prim-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{$\rho, S, H, (+~x~y), \sigma$ \\ $\rightsquigarrow \rho', (\rho, x, (+~x~y), \{\psi_{i}\}, \sigma)\!:\!S, H, s, \bigcup_{\psi \in \{\psi_{i}\}}\mmap{}{\psi}{x}$}      &
\sc{prim-1-clo} \\
\hline
$H(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & \makecell{$\rho, S, H, (+~x~y), \sigma$ \\ $\rightsquigarrow \rho', (\rho, y, (+~x~y), \{\psi_{i}\}, \sigma)\!:\!S, H, s, \bigcup_{\psi \in \{\psi_{i}\}}\mmap{}{\psi}{y} $}      &
\sc{prim-2-clo} \\
\hline
{$\mathit{f}~\mbox{defined as}$
$~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & \makecell{$\rho, S, H, (f~\myvec{x}), \sigma$ \\ $\rightsquigarrow [\myvec{y} \mapsto \rho(\myvec{x})], S, H, e_{\mathit{f}}, \sigma$}      &
\sc{funcall} \\
\hline
$\ell$ is a new location& \makecell{$\rho, S, H, (\LET~x\leftarrow s~\IN~e)$ \\ $ \rightsquigarrow \rho\oplus[x \mapsto \ell], S, H[\ell \mapsto \langle s, \lfloor\rho\rfloor_{FV(s)}  \oplus [x \mapsto
  \ell]\rangle], e$} &
\sc{let} \\
\hline
$H(\rho(x)) \ne 0$ & \makecell{$\rho, S, H, (\pi:\SIF~\psi:x~e_1~e_2), ep, \sigma$ \\ $\rightsquigarrow \rho, S, H,  e_1, ep, \sigma$} & \sc{if-true} \\
\hline
$H(\rho(x)) = 0$ & \makecell{$\rho, S, H, (\pi:\SIF~\psi:x~e_1~e_2), ep, \sigma$ \\   $\rightsquigarrow
\rho, S, H,  e_2, ep, \sigma$} & \sc{if-false} \\
\hline
$H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H,  (\pi:\SIF~\psi:x~e_1~e_2), \sigma$ \\ $\rightsquigarrow
\rho', (\rho, x, (\SIF~x~e_1~e_2),  \pmap{}(\pi), \sigma)\!:\!S, H,  s, \mmap{}{\psi}{x}$}
&
\sc{if-clo} \\
\hline
{$H(\rho(x))~\mbox{is}$ $\mbox{whnf with value}~v$}& \makecell{$\rho, (\rho', z, e, ep, \sigma')\!:\!S, H, (\SRETURN~x), \sigma$ \\ $\rightsquigarrow \rho', S, H[\rho(z) \mapsto v], e, \sigma'$} &
\sc{return-whnf}\\
\hline
$H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H, (\psi:\SRETURN~x), \sigma$ \\ $
  \rightsquigarrow$
$\rho',~ (\rho, x, (\SRETURN~x), \{\psi\}, \sigma)\!:\!S, H,  s, \mmap{}{\psi}{x}$} &
\sc{return-clo} \\
\hline
\end{tabular}
\caption{Minefield semantics for the language.\label{fig:minefield-semantics}}
\end{center}
\end{figure*}

\bigskip

\begin{itemize}
\item State of the operational semantics will be $\rho, S, H, e,
  \sigma$.
\item Stack content will be $\rho, x, e, \sigma, eps$.
\item Liveness environment calculation:
  \begin{enumerate}
  \item Of the   current expression being
  evaluated:\\  
  $ \Lv = \bigcup_{ep \in \mathcal{P}(e)} \mathcal{M}_\sigma(ep)$, where $
  \mathcal{M}_\sigma = \mathcal{L}(e, \sigma, \Lfonly)$
\item Liveness environment of a stacked computation\\
  $ \Lv = \bigcup_{ep \in eps} \mathcal{M}_\sigma(ep), \text{where }
  \mathcal{M}_\sigma = \mathcal{L}(e, \sigma, \Lfonly)$ 
  \end{enumerate}
\item $GC$ models a liveness based garbage collection:
$GC(\rho, S, H, L, SL) = (\rho', S', H')$ 
  \begin{enumerate}
  \item For each $x \in domain(\rho)$,  $\rho'(x)=\bot$ iff $x.\epsilon
    \notin \Lv$.
\item For each $x \in domain(\rho)$ where $(\rho,\_,\_,\_,\_,\_)$ is a
  stack entry  with \Lv\ as the  liveness environment, $\rho'(x)=\bot$
  iff $x.\epsilon \notin \Lv$.
\item for each location $\ell$, $H'(\ell) = \bot$, iff  there is no
  $x$ in either the current environment or one of the stacked
  environments $\rho$ such that for some $\alpha$, $H[x.\alpha] =
  \ell$  and $x.\alpha \notin \Lv$, .

  \end{enumerate}
\end{itemize}
